Missing the Point\(\colon\) Non-Convergence in Iterative Imputation Procedures

### 2020-04-23 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Both Rhat and AC work properly now for all iterations. Last version with old functions. Cleaning up Functions folder after this.

\textbf{Note for peer review: things written in bold are reminders for
myself or things that I am unsure off still. Also, I just decided to
change things majorly in my results section. So right now, the figures
do not correspond to the text. Please only skim that section, because I
am definitely going to rewrite it. Thanks already! - Hanne}

@@

Let \(y\) denote an \(n \times p\) matrix containing the data values on
\(p\) variables for all \(n\) units in a sample. The data value of unit
\(i\) (\(i = 1, \dots, n\)) on variable \(j\) (\(j = 1, \dots, p\)) may
be either observed or missing. The collection of observed data values in
\(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred
to as \(y_{mis}\). For each datapoint in \(y_{mis}\), we sample
\(M \times T\) times plausible values, where \(M\) is the number of
imputations (\(m = 1, 2, \dots, M\)) and \(T\) is the number of
iterations (\(t = 1, 2, \dots, T\)). The collection of samples between
the initial value (at \(t=1\)) and the final imputed value (at \(t=T\))
will be referred to as an `imputation chain'. \textbf{Add: missingness
percentage = proportion of cases with one or more missing values, Q is
quantity of scientific interest: ``A scientific estimand \(Q\) is a
quantity of scientific interest that we can calculate if we would
observe the entire population'' \citep[par 2.3.1]{buur18}. }

@@

The two challenges of convergence are stationarity and mixing point.
Mixing and stationarity can be evaluated visually. Common practice is to
evaluate stationarity and mixing in trace plots for scalar summaries of
interest. In mice these scalar summaries of interest are chain means and
chain variances. As \citet{buur18} describes, users can also specify a
model-specific scalar summary (e.g., a regression coefficient). Such a
user-specified scalar summary, however, is not universal to all complete
data problems. Therefore, as inspired by \textbf{someone} (2002), we
will use introduce a scalar summary that summarizes the multivariate
state space of the algorithm. In this study we use the first eigenvalue
of the variance covariance matrix of the completed data set per
imputation. The methods under evaluation are \(\widehat{R}\) and \(AC\).
These will be computed for the three scalar summaries of interest: chain
means, chain variances, and the first eigenvalue of the
variance-covariance matrix across imputations.

The two methods will be evaluated with an estimand per scalar summary of
interest. It is assumed that when the estimates are unbiased (and
confidence valid), the algorithm is sufficiently converged. The
\(\widehat{R}\) and \(AC\) values with chain means as theta are
evaluated against the bias in univariate mean estimates (\textbf{just
for Y??}). Similarly, the performance measure for Rhat and \(AC\)
applied to chain variances is the bias in estimated standard deviations.
To evaluate the performance of Rhat and \(AC\) on the eigenvalues, we
will use bias in R squared and in the estimated regression coefficient,
the coverage rate of the CI95\% of the regression estimate, and the CI
length. \textbf{remove the term performance measure above and add that
performance measures are bias in all estimands, and cov and CI length of
regression coeff.}

\textbf{Result:} In general there is more bias in the estimate in
conditions with a higher proportion of missingness and or a lower number
of iterations. As required the values of \(\widehat{R}\) are lower for
conditions with a higher number off iterations. End somewhat higher in
conditions with a higher percentage of missing data. However
autocorrelation did not show a decrease with increase number of
iterations. After replacing the ACF function with a manual calculation
for autocorrelation the autocorrelation values were indeed decreasing
with a higher number of iterations. Evaluation with the performance
measures shows that are had an autocorrelation are conservative: signs
of non-convergence in conditions where estimands are unbiased and or
confidence valid. As expected complete convergence (as indicated by
autocorrelation equals 0; \(\widehat{R}\) equals 1) is not observed.

@@

Usually, convergence is diagnosed when Rhat \textless{} 1.2 or 1.1 or
1.01. And a t-test is performed for \(AC\) values == 0. Performance
measures are the same as above: unbiased, confidence valid estimates.

\textbf{Result:} \(\widehat{R}\) is equal to 1.2 is not stringent enough
because the performance measures R not unbiased and confidence fellas
fell it 4 every condition in which \(\widehat{R}\) is smaller than one
point 2. At least, not when \(\widehat{R}\) is smaller than one point
two fur first occurs. The threshold \(\widehat{R}\) is equal to one
point one is somewhat conservative with respect to the performance
measures. It seems OK. \(\widehat{R}\) is equal to 1.01 is too stringent
compared to the performance measures (\textbf{but may be necessary in
more complex complete data models???}).

\hypertarget{start-methods-section}{%
\subsection{Start methods section}\label{start-methods-section}}

@@

\hypertarget{results}{%
\section{Results}\label{results}}

@@

When applied to the imputation chain means, \(\widehat{R}\) indicates
that the mice algorithm does not reach a converged state
(\(\widehat{R}\) = 1) in any of the simulation conditions. Neither is
the most recent recommended threshold reached (\(\widehat{R}\)
\textless{} 1.01). The conventional \(\widehat{R}\) threshold of 1.2 is
reached in simulation conditions \(T = 3\) and \(T > 6\). (With the
default number of iterations (maxit = 5), this dip in Rhat values would
be spotted, so it is no problem.) The point at which an extra iteration
does not seem to improve the Rhat value is around \(T=30\).

@@

If I don't include it in my study, also: Significance of \(AC\) values;
Estimate (PCA or reg. coeff.) versus Rhat and \(AC\) on the estimate

>>>>

\textbf{Dear Daniela, I have just implemented some major changes in my
Methods and Results sections. Therefore, these are kind of a mess.
Please do not waste any time on spelling/wording there, I will continue
improving it. On the other hand, if you have general advice, please tell
me because I'm happy to move thngs around still! Figures in the Results
are currently missing, but will help in understanding the paper. Sorry
that you won't have that now! Also, things written in bold are reminders
for myself or things that I am unsure off still. Thanks already! -
Hanne}

@@

Let \(y\) denote an \(n \times p\) matrix containing the data values on
\(p\) variables for all \(n\) units in a sample. The data value of unit
\(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\)
(\(j = 1, 2, \dots, p\)) may be either observed or missing. The
collection of observed data values in \(y\) is denoted by \(y_{obs}\);
the missing part of \(y\) is referred to as \(y_{mis}\). For each
datapoint in \(y_{mis}\), we sample \(M \times T\) times plausible
values, where \(M\) is the number of imputations
(\(m = 1, 2, \dots, M\)) and \(T\) is the number of iterations
(\(t = 1, 2, \dots, T\)). The collection of samples between the initial
value (at \(t=1\)) and the final imputed value (at \(t=T\)) will be
referred to as an `imputation chain'. \textbf{Add: missingness
percentage = proportion of cases with one or more missing values, Q is
quantity of scientific interest: ``A scientific estimand \(Q\) is a
quantity of scientific interest that we can calculate if we would
observe the entire population'' \citep[par 2.3.1]{buur18}, and
\(\theta\)s are scalar summaries of interest in the iterative algorithm
(e.g., chain means; the average of the imputed values in each imputation
chain). }

@@

The two challenges of convergence are stationarity and mixing. Mixing
and stationarity can be inspected visually, by evaluating traceplots of
scalar summaries of interest (\(\theta\)s; e.g., chain means and chain
variances). As \citet{buur18} describes, users can also specify a
model-specific scalar summary (e.g., a regression coefficient). A
user-specified scalar summary, however, is not universal to all complete
data problems. Therefore, as inspired by \textbf{someone} (2002), we
will use introduce a \(\theta\) that summarizes the multivariate state
space of the algorithm. Namely, the first eigenvalue of the
variance-covariance matrix of the \(M\) completed datasets.

The methods under evaluation are \(\widehat{R}\) and \(AC\). These will
be computed for the three scalar summaries of interest: chain means
(i.e., per imputation, per variable: on \(y_{imp, m, j}\)), chain
variances (i.e., per imputation, per variable: on \(y_{imp, m, j}\)),
and the first eigenvalue of the variance-covariance matrix across
imputations (i.e., per imputation: on \{\(y_{obs}, y_{imp, m}\)\})
\textbf{add user-specified \(\theta\): estimated regression coefficient
per imputation}.

The performance of two methods (\(\widehat{R}\) and \(AC\)) on each of
these scalar summaries of interest will be evaluated with several
estimands (quantities of scientific interest; \(Q\)s). It is assumed
that when \(\bar{Q}\) (the pooled result across imputations) is an
unbiased (and confidence valid) estimate of \(Q\), the algorithm is
sufficiently converged. For each scalar \(\theta\), one estimand \(Q\)
is defined that may be of interest in empirical research. The
\(\widehat{R}\) and \(AC\) values with chain means as \(\theta\)s are
evaluated against the bias in univariate mean estimates (\textbf{all
variables or just for Y??}). Similarly, the performance measure for
\(\widehat{R}\) and \(AC\) applied to chain variances is the bias in
estimated standard deviations. To evaluate the performance of
\(\widehat{R}\) and \(AC\) on the eigenvalues, we will use bias in the
coefficient of determination, \(R^2\). Additionally, we will evaluate
the estimated regression coefficients, the coverage rate of the CI95\%
of the regression estimates, and the CI length (\textbf{see definitions
in old Methods}). \textbf{remove the term performance measure here and
add an extra alinea to define performance measures as bias in all
estimands, and coverage rate and CI length of regression coefficients.}

\textbf{Move this to Results:} As expected, conditions with a higher
proportion of missingness and/or a lower number of iterations show more
signs of non-convergence, as indicated by more extreme bias in the
estimated \(Q\)s. Roughly speaking this means that MICE is indeed not
converged as \(t=1\), and converges gradually as \(t\) increases. The
point at wich an additional iteration does not lead to an improvement of
the estimates depends on the difficulty of the missingness
problem---between XYZ and XYZ. This means that the algorithm has
converged sufficiently (\textbf{under the current specifications}).

As required for a method to diagnose non-convergence, the values of
\(\widehat{R}\) follow a trend similar to the performance measures.
\(\widehat{R}\) values are generally lower in conditions with a higher
number of iterations, and somewhat higher in conditions with a higher
percentage of missing data. \textbf{Decide where to specify the two
calculations of AC:} Initially, autocorrelation did not show a decrease
with an increasing number of iterations. But, after replacing the ACF
function with a manual calculation for autocorrelation, the
autocorrelation values were indeed decreasing with a higher number of
iterations.

Evaluation with the performance measures shows that are had an
autocorrelation are conservative: they indicate signs of non-convergence
in conditions where \(\bar{Q}\)s are unbiased and or confidence valid
estimates of \(Q\).

@@

Upon convergence, \(\widehat{R}=1\) and \(AC=0\), but in MCMC algorithms
these are very unlikely (because it's a distribution, not a point to
converge to). In practice thresholds are used to diagnose
non-convergence. Usually, non-convergence is diagnosed when
\(\widehat{R}\) \textgreater{} 1.2 or 1.1 or even 1.01. And a t-test is
performed to assess whether \(AC\) is significantly different from zero.
Performance measures are the same as above: unbiased, confidence valid
estimates.

\textbf{Move this to Results:} As expected, complete convergence
(\(\widehat{R}=1\) and \(AC=0\)) is not observed. But what about the
thresholds for practice? \(\widehat{R}\) \textless1.2 is not stringent
enough, because conditions in which \(\widehat{R}\) is smaller than 1.2
do not all yield unbiased estimates of the \(Q\)s. Moreover, there is a
dip in the \(\widehat{R}\) values, after which \(\widehat{R}\)
\textgreater{} 1.2 again. If the algorithms would be terminated at the
iteration where \(\widehat{R}\) \textless{} 1.2 occurs for the first
time, the increase in \(\widehat{R}\) values might be missed. The
threshold \(\widehat{R}\) \textless{} 1.1 is somewhat conservative with
respect to the performance measures. But it seems OK. \(\widehat{R}\)
\textless{} 1.01 is too stringent compared to the performance measures.
And it is not obtained for any number of iterations (\textbf{but may be
necessary in more complex complete data models???}).

\hypertarget{start-old-methods-section}{%
\subsection{Start old methods section}\label{start-old-methods-section}}

@@

\hypertarget{old-results-section}{%
\section{Old Results section}\label{old-results-section}}

@@

When applied to the imputation chain means, \(\widehat{R}\) indicates
that the mice algorithm does not reach a converged state
(\(\widehat{R}\) = 1) in any of the simulation conditions. Neither is
the most recent recommended threshold reached (\(\widehat{R}\)
\textless{} 1.01). The conventional \(\widehat{R}\) threshold of 1.2 is
reached in simulation conditions \(T = 3\) and \(T > 6\). (With the
default number of iterations (maxit = 5), this dip in \(\widehat{R}\)
values would be spotted, so it is no problem.) The point at which an
extra iteration does not seem to improve the \(\widehat{R}\) value is
around \(T=30\).

@@

If I don't include it in my study, also: Significance of \(AC\) values;
Estimate (PCA or reg. coeff.) versus \(\widehat{R}\) and \(AC\) on the
estimate
